---
title: "TransferMarket"
author: "Siddiq Khan, Morgan Fissel, Raunav Sharma"
date: "2023-11-30"
output: html_document
---

```{r}
library(ggplot2)
library(scales)
library(dplyr)
library(mgcv)
library(maps)
library(mapdata)
```



```{r}

transfer_data <- read.csv("merged_cleaned_data.csv")
head(transfer_data)


```


```{r}

str(transfer_data)
```

```{r}

unique_positions <- unique(transfer_data$short_pos)
print(unique_positions)

```



```{r}

transfer_data <- transfer_data %>%
  mutate(position_category = case_when(
    short_pos %in% c("CM", "LM", "AM", "DM", "RM") ~ "midfield",
    short_pos %in% c("RB", "LB", "CB") ~ "defence",
    short_pos %in% c("SS", "LW", "RW", "CF") ~ "attack",
    TRUE ~ short_pos  
  ))

head(transfer_data)

```

```{r}

unique_positions <- unique(transfer_data$position_category)
print(unique_positions)

```


```{r}
filtered_data <- transfer_data[transfer_data$fee > 0, ]
```



```{r}
transfer_data_2021 <- transfer_data[transfer_data$season == "2021" & transfer_data$fee > 0, ]

ggplot(transfer_data_2021, aes(x = market_value, y = fee, color = league)) +
  geom_point() +
  labs(title = "Relationship between Market Value and Transfer Fee (All Leagues - 2021)",
       x = "Market Value",
       y = "Transfer Fee",
       color = "League")
```
```{r}
ggplot(transfer_data, aes(x = age, fill = ..count..)) +
  geom_histogram(binwidth = 1, color = "black", alpha = 0.7) +
  scale_fill_gradient("Count", low = "yellow", high = "purple") +
  labs(title = "Transfer Distribution by Age from 2011 to 2020",
       x = "Age",
       y = "Count") +
  theme_minimal()

ggplot(transfer_data, aes(x = age, y = fee, color = league)) +
  geom_point(alpha = 0.7) +
  #scale_color_manual(values = league_colors, name = "League") +
  labs(title = "Transfer Fees versus Age",
       x = "Age",
       y = "Transfer Fee",
       color = "League") +
  theme_minimal()

```


```{r}
# x_limits =c(0,12500000)
# y_limits =c(0,15000000)
## faceted by league

ggplot(filtered_data, aes(x = market_value, y = fee)) +
  geom_point() +
  labs(title = "Relationship between Market Value and Transfer Fee (All Leagues)",
       x = "Market Value",
       y = "Transfer Fee") + facet_wrap(~ league)

ggplot(filtered_data, aes(x = log(market_value), y = log(fee))) +
  geom_point() +
  labs(title = "Relationship between log(Market Value) and log(Transfer Fee) (All Leagues)",
       x = "Market Value",
       y = "Transfer Fee") +
  facet_wrap(~ league, scales = "free")  

```

```{r}
ggplot(filtered_data, aes(x = market_value, y = fee, color = position_category)) +
  geom_point(aes(group = position_category), size = 1) +
  #geom_point(size = 1) +
  labs(
    title = "Scatter Plot - Market Value and Position on Transfer Fees",
    x = "Market Value",
    y = "Transfer Fee",
    color = "Position"
  ) + 
  facet_wrap(~position_category)

##Log Scale
ggplot(filtered_data, aes(x = log(market_value), y = log(fee), color = position_category)) +
  geom_point(aes(group = position_category), size = 1) +
  #geom_point(size = 1) +
  labs(
    title = "Scatter Plot - log(Market Value) and Position on log(Transfer Fees)",
    x = "Market Value",
    y = "Transfer Fee",
    color = "Position"
  ) + 
  facet_wrap(~position_category)
```






```{r}
# Filter the data for the year 2021 and exclude players with a fee of 0
transfer_data_2021 <- transfer_data[transfer_data$season == "2021" & transfer_data$fee > 0, ]

# Scatter plot with color-coded points for each league in the year 2021
ggplot(transfer_data, aes(x = market_value, y = fee, color = league)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +  # Add a linear regression line
  labs(title = "Relationship between Market Value and Transfer Fee (All Leagues)",
       x = "Market Value",
       y = "Transfer Fee",
       color = "League") + 
  facet_grid(league ~ .)

```

```{r}
# Create a new variable for year categories
filtered_data$year_category <- cut(filtered_data$season, breaks = c(2011, 2014, 2018, 2021), labels = c("2011-2014", "2015-2018", "2019-2021"))

# Filter out rows with NA in year_category
filtered_data <- filtered_data[!is.na(filtered_data$year_category), ]

# Visualize the scatter plot with categorized years
ggplot(filtered_data, aes(x = market_value, y = fee, color = league)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue", formula = y ~ x) +
  labs(title = "Market Value vs Transfer Fees - Categorized Years",
       x = "Market Value",
       y = "Transfer Fee",
       color = "League") +
  facet_grid(league ~ year_category)

```





```{r}
# Filter the data for the year 2021 and exclude players with a fee of 0
transfer_data_2021 <- transfer_data[transfer_data$season == "2021" & transfer_data$fee > 0, ]

# Scatter plot with color-coded points for each league in the year 2021
ggplot(transfer_data_2021, aes(x = market_value, y = fee, color = league)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +  # Add a linear regression line
  labs(title = "Relationship between Market Value and Transfer Fee (All League) 2021",
       x = "Market Value",
       y = "Transfer Fee",
       color = "League") + 
  facet_grid(league ~ .)

```



```{r}
transfer_data_2021 <- transfer_data[transfer_data$season == "2021" & transfer_data$fee > 0, ]

# Define unique colors for each league
league_colors <- c("Premier League" = "red", "Laliga" = "blue", "1 Bundesliga" = "green", "Serie A" = "purple", "Ligue 1" = "orange")

# Scatter plot with color-coded points and separate trend lines for each league
ggplot(transfer_data_2021, aes(x = market_value, y = fee, color = league)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, aes(group = league, color = league)) +  # Separate trend lines for each league with different colors
  scale_color_manual(values = league_colors, name = "League") +  # Set colors according to league
  labs(title = "Relationship between Market Value and Transfer Fee (All Leagues - 2021)",
       x = "Market Value",
       y = "Transfer Fee",
       color = "League")
```

```{r}


create_scatter_plot <- function(data, title) {
  ggplot(data, aes(x = market_value, y = fee, color = league)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE, aes(group = league, color = league)) +
    scale_color_manual(values = league_colors, name = "League") +
    labs(title = title,
         x = "Market Value",
         y = "Transfer Fee",
         color = "League")
}

transfer_data_2021 <- transfer_data[transfer_data$season == "2021" & transfer_data$fee > 0, ]
plot_2021 <- create_scatter_plot(transfer_data_2021, "Relationship between Market Value and Transfer Fee (All Leagues - 2021)")

transfer_data_2016 <- transfer_data[transfer_data$season == "2016" & transfer_data$fee > 0, ]
plot_2016 <- create_scatter_plot(transfer_data_2016, "Relationship between Market Value and Transfer Fee (All Leagues - 2016)")

transfer_data_2011 <- transfer_data[transfer_data$season == "2011" & transfer_data$fee > 0, ]
plot_2011 <- create_scatter_plot(transfer_data_2011, "Relationship between Market Value and Transfer Fee (All Leagues - 2011)")

par(mfrow = c(1, 3))

plot_2021
plot_2016
plot_2011


```


```{r}

# Calculate the correlation coefficient
correlation_coefficient <- cor(filtered_data$market_value, filtered_data$fee)

# Print the correlation coefficient
cat("Correlation Coefficient:", correlation_coefficient, "\n")
```



Using filtered_data:
```{r}
model <- lm(fee ~ market_value + league, data = filtered_data)

# Plot
ggplot(filtered_data, aes(x = market_value, y = fee, color = league)) +
  geom_point() +
  geom_abline(intercept = coef(model)[1], slope = coef(model)[2], color = "blue") +
  labs(title = "Relationship between Market Value and Transfer Fee (All League)",
       x = "Market Value",
       y = "Transfer Fee",
       color = "League") + 
  facet_grid(league ~ .)
```


```{r}
model2 <- lm(fee ~ market_value + short_pos, data = filtered_data)

# Plot
ggplot(filtered_data, aes(x = market_value, y = fee, color = position_category)) +
  geom_point() +
  geom_abline(intercept = coef(model)[1], slope = coef(model)[2], color = "blue") +
  labs(title = "Relationship between Market Value and Transfer Fee (All League)",
       x = "Market Value",
       y = "Transfer Fee",
       color = "League") + 
  facet_grid(league ~ .)
```





```{r}
model1 <- lm(fee ~ market_value + short_pos + league + season, data = filtered_data)

ggplot(filtered_data, aes(x = market_value, y = fee, color = league)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue", formula = y ~ x) +  # Add a linear regression line
  labs(title = "Relationship between Market Value and Transfer Fee (Model)",
       x = "Market Value",
       y = "Transfer Fee",
       color = "League") +
  facet_grid(league ~ .)

```

############################################################
MODEL WE CHOSE 

Linear Regression with Age Categories
```{r}

# Create age categories
filtered_data$age_category <- cut(filtered_data$age, breaks = c(18, 21, 25, 30, 35, 40, 45, 50, 100),
                                   labels = c("18-21", "22-25", "26-30", "31-35", "36-40", "41-45", "46-50", "50+"))

# Fit a linear regression model with age categories
model_age_cat <- lm(fee ~ market_value + short_pos + league + season + as.factor(age_category), data = filtered_data)

# Visualize the relationship
ggplot(filtered_data, aes(x = market_value, y = fee, color = league)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue", formula = y ~ x) +
  labs(title = "Relationship between Market Value and Transfer Fee (Model with Age Categories)",
       x = "Market Value",
       y = "Transfer Fee",
       color = "League") + 
  facet_grid(league ~ .)


```

ANALYSIS OF CHOSEN MODEL 
```{r}
library(car)
vif(model_age_cat)

```


```{r}
# Add predictions and actual values to the dataset
filtered_data$predictions_age_cat <- predict(model_age_cat, newdata = filtered_data)
filtered_data$actual_values <- filtered_data$fee

# Visualize predictions vs. actual values
ggplot(filtered_data, aes(x = actual_values, y = predictions_age_cat, color = league)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  labs(title = "Predictions vs. Actual Transfer Fees (Model with Age Categories)",
       x = "Actual Transfer Fee",
       y = "Predicted Transfer Fee",
       color = "League")


```



```{r}

residuals_age_cat <- residuals(model_age_cat)

# Assuming log transformation is needed for market_value
filtered_data$log_market_value <- log(filtered_data$market_value)

# Ensure both data frames have the same number of rows
min_rows <- min(nrow(filtered_data), length(residuals_age_cat))
filtered_data <- head(filtered_data, min_rows)
residuals_data <- data.frame(
  log_market_value = filtered_data$log_market_value,
  residuals = head(residuals_age_cat, min_rows),
  league = filtered_data$league
)

# Visualize residuals
ggplot(residuals_data, aes(x = log_market_value, y = residuals, color = league)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residuals vs. Log-transformed Market Value (Model with Age Categories)",
       x = "Log-transformed Market Value",
       y = "Residuals",
       color = "League") +
  facet_grid(league ~ .)




```







```{r}
# Predict transfer fees
predictions_age_cat <- predict(model_age_cat, newdata = filtered_data)

# Evaluate accuracy
accuracy <- sqrt(mean((filtered_data$fee - predictions_age_cat)^2))
print(paste("Root Mean Squared Error (RMSE):", round(accuracy, 2)))

```


```{r}
# Check for missing values
any(is.na(filtered_data$fee))
any(is.na(predictions_age_cat))

```




```{r}
# Check unique values and missing values in age_category
table(filtered_data$age_category, useNA = "always")

```


```{r}
# Remove rows with missing age_category
filtered_data <- na.omit(filtered_data, cols = "age_category")

```


```{r}
# Predict transfer fees
filtered_data$predictions_age_cat <- predict(model_age_cat, newdata = filtered_data)

# Evaluate accuracy
accuracy <- sqrt(mean((filtered_data$fee - filtered_data$predictions_age_cat)^2))
print(paste("Root Mean Squared Error (RMSE):", round(accuracy, 2)))

```



ANALYSIS USING PREDICTED VALUES

Positional Analysis using position_category:
```{r}
# Create a summary table for average predicted transfer fees by position category
position_category_summary <- aggregate(predictions_age_cat ~ position_category, data = filtered_data, FUN = mean)


```


Comparative Analysis for Position Categories:
```{r}
library(ggplot2)
ggplot(filtered_data, aes(x = position_category, y = predictions_age_cat, fill = position_category)) +
  geom_boxplot() +
  labs(title = "Comparison of Predicted Transfer Fees by Position Category",
       x = "Position Category",
       y = "Predicted Transfer Fee",
       fill = "Position Category")


```


Positional Trends Over Time by Category:
```{r}
ggplot(filtered_data, aes(x = season, y = predictions_age_cat, color = position_category)) +
  geom_line() +
  labs(title = "Position Category Trends in Predicted Transfer Fees Over Time",
       x = "Season",
       y = "Predicted Transfer Fee",
       color = "Position Category")

```



Positional Influence in Different Leagues by Category:
```{r}
ggplot(filtered_data, aes(x = position_category, y = predictions_age_cat, color = league)) +
  geom_point() +
  facet_wrap(~league) +
  labs(title = "Impact of Position Category on Predicted Transfer Fees by League",
       x = "Position Category",
       y = "Predicted Transfer Fee",
       color = "League")

```



Top Value Position Categories:
```{r}
# Identify top position categories by average predicted transfer fees
top_position_categories <- position_category_summary[order(position_category_summary$predictions, decreasing = TRUE), ]

top_position_categories
```






Club Analysis using Club:
```{r}
# Create a summary table for average predicted transfer fees by club
club_summary <- aggregate(predictions_age_cat ~ club, data = filtered_data, FUN = mean)

```


Top Value Clubs: which football clubs, on average, tend to have higher predicted transfer fees:
```{r}
# Identify top clubs by average predicted transfer fees
top_clubs <- club_summary[order(club_summary$predictions, decreasing = TRUE), ]
top_clubs

```




Nationality Analysis using Nationality:
```{r}
# Create a summary table for average predicted transfer fees by nationality
nationality_summary <- aggregate(predictions_age_cat ~ nationality, data = filtered_data, FUN = mean)

```



```{r}
library(ggplot2)

# Filter the top 10 nationalities
top_nationalities <- nationality_summary %>%
  top_n(10, wt = predictions_age_cat)

# Bar chart for the top 10 nationalities
ggplot(top_nationalities, aes(x = reorder(nationality, predictions_age_cat), y = predictions_age_cat)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Top 10 Nationalities - Predicted Transfer Fees",
       x = "Nationality",
       y = "Predicted Transfer Fee")



# World map 

# Create a world map
world_map <- map_data("world")

library(maps)
ggplot(filtered_data, aes(map_id = nationality)) +
  geom_map(aes(fill = predictions_age_cat), map = world_map) +
  expand_limits(x = world_map$long, y = world_map$lat) +
  theme_void() +
  labs(title = "Predicted Transfer Fees by Nationality on World Map",
       fill = "Predicted Transfer Fee")

```



```{r}
# Identify the top nationalities by average predicted transfer fees
top_nationalities <- head(nationality_summary[order(-nationality_summary$predictions), ], 10)

# Print the top nationalities
print(top_nationalities)

```






#############################################################

Quantifying Difference Between Leagues Over Time
```{r}
# Linear regression model with league, season, and league:season interaction
model3 <- lm(fee ~ market_value + short_pos + league * season, data = filtered_data)

# Visualize the league differences over time
ggplot(filtered_data, aes(x = season, y = fee, color = league)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue", formula = y ~ x) +
  labs(title = "Difference in Transfer Fees Between Leagues Over Time",
       x = "Season",
       y = "Transfer Fee",
       color = "League") +
  facet_grid(league ~ .)


```



Adding Interaction Terms
```{r}
# Model with interaction terms
model_interaction <- lm(fee ~ market_value * short_pos * league * season * as.factor(age_category), data = filtered_data)

# Visualization
ggplot(filtered_data, aes(x = market_value, y = fee, color = league)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue", formula = y ~ x) +  
  labs(title = "Model with Interaction Terms",
       x = "Market Value",
       y = "Transfer Fee",
       color = "League") +
  facet_grid(league ~ .)


```




Interaction Model - league:season interaction and log-transformed variables
```{r}
# Log-transform fee and market value
filtered_data$log_fee <- log(filtered_data$fee)
filtered_data$log_market_value <- log(filtered_data$market_value)

# Model with league:season interaction and log-transformed variables
interaction_model <- lm(log_fee ~ log_market_value + short_pos + league * season, data = filtered_data)

# Visualization
ggplot(filtered_data, aes(x = log_market_value, y = log_fee, color = league)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue", formula = y ~ x) +
  labs(title = "Relationship between Log-transformed Market Value and Log-transformed Transfer Fee",
       x = "Log-transformed Market Value",
       y = "Log-transformed Transfer Fee",
       color = "League") +
  facet_grid(league ~ .)



```




```{r}

# Check model summary
summary(interaction_model)

# Check variance inflation factor (VIF) for multicollinearity
library(car)
vif(interaction_model)

# Add predictions and actual values to the dataset
filtered_data$predictions <- exp(predict(interaction_model))
filtered_data$actual_values <- filtered_data$fee

# Visualize predictions vs. actual values
ggplot(filtered_data, aes(x = actual_values, y = predictions, color = league)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  labs(title = "Predictions vs. Actual Transfer Fees",
       x = "Actual Transfer Fee",
       y = "Predicted Transfer Fee",
       color = "League")

# Check residuals
residuals <- residuals(interaction_model)

# Visualize residuals
ggplot(filtered_data, aes(x = log_market_value, y = residuals, color = league)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residuals vs. Log-transformed Market Value",
       x = "Log-transformed Market Value",
       y = "Residuals",
       color = "League") +
  facet_grid(league ~ .)



```


```{r}
# Evaluate model performance
library(Metrics)
rmse_value <- rmse(filtered_data$actual_values, filtered_data$predictions)
cat("Root Mean Squared Error (RMSE):", rmse_value, "\n")



```







AICs
```{r}
model <- lm(fee ~ market_value + league, data = filtered_data)
model1 <- lm(fee ~ market_value + short_pos + league + season, data = filtered_data)
model2 <- lm(fee ~ market_value + short_pos, data = filtered_data)
model3 <- lm(fee ~ market_value + short_pos + league * season, data = filtered_data)
model_age_cat <- lm(fee ~ market_value + short_pos + league + season + as.factor(age_category), data = filtered_data)
model_interaction <- lm(fee ~ market_value * short_pos * league * season * as.factor(age_category), data = filtered_data)
interaction_model <- lm(log_fee ~ log_market_value + short_pos + league * season, data = filtered_data)




# AIC values
AIC(model, model1, model2, model3, model_age_cat, model_interaction, interaction_model)

```

```{r}
correlation_coefficient <- cor(filtered_data$market_value, filtered_data$season)

# Print the correlation coefficient
cat("Correlation Coefficient:", correlation_coefficient, "\n")
#model 1 is the lowest
# if i were to suspect interaction it would be between position and season
model1 <- lm(fee ~ market_value + position_category + league + season, data = filtered_data)
lm_all <- lm(fee ~ market_value + position_category + league + season + age, data = filtered_data)
lm_age_val <- lm(fee ~ market_value*age + position_category + league + season, data = filtered_data)
lm_age_val2 <- lm(fee ~ market_value*age + position_category + league*season, data = filtered_data)

AIC(model1, lm_all, lm_age_val,lm_age_val2)


```
```{r}


ggplot(filtered_data, aes(x = market_value, y = fee, color = position_category)) +
  geom_point() +
  geom_abline(intercept = coef(lm_all)[1], slope = coef(lm_all)[2], color = "purple") +
  labs(title = "Relationship between Market Value and Transfer Fee (All League)",
       x = "Market Value",
       y = "Transfer Fee",
       color = "League") + 
  facet_grid(league ~ .)
```


```{r}
lm_age_val <- lm(fee ~ market_value*age + position_category + league + season, data = filtered_data)


ggplot(filtered_data, aes(x = market_value, y = fee, color = position_category)) +
  geom_point() +
  geom_abline(intercept = coef(lm_age_val)[1], slope = coef(lm_age_val)[2], color = "purple") +
  labs(title = "Relationship between Market Value and Transfer Fee (All League)",
       x = "Market Value",
       y = "Transfer Fee",
       color = "League") + 
  facet_grid(league ~ .)

```

```{r}
# Create a GAM model with a nonlinear term in age and market_value
gam_model <- gam(fee ~ s(market_value) + s(age) + position_category + league + s(season),
                 data = filtered_data)


# Plot the GAM model
plot(gam_model, select = 2)  # Select the age smooth term for plotting

# Summary of the GAM model
summary(gam_model)

filtered_data$predicted_fee <- predict(gam_model, newdata = filtered_data)

# Plot the observed data and the fitted values
ggplot(filtered_data, aes(x = market_value, y = fee, color = league)) +
  geom_point(alpha = 1) +  # Scatter plot of observed data
  geom_line(aes(y = predicted_fee), color = "purple", size = 1) +  # Fitted values from the GAM model
  labs(title = "Observed Data vs. Fitted Values (GAM Model)",
       x = "Market Value",
       y = "Transfer Fee")

```

```{r}


```

```{r}


```



```{r}


  
```
